E-Graphs With Bindings

### 2025-01-24 LICS_2025/e_graphs_with_bindings/sections/categorical_semantics.tex
sections done

Given a category $\mathbb{C}$  with objects $A,B \in \mathbb{C}$ we denote by $\mathbb{C}(A,B)$ the corresponding hom-set.  
We write the identity morphism on $A$ as $\id_A$.  
We commonly write $f;g$ for composition in diagrammatic order.  
Composition in the usual order is written $g \circ f$.  
We denote the tensor product of an SMC $\mathbb{C}$ by $\otimes$,  its unit by $I$, and its symmetry natural transformation as $\sym$ \cite{maclane}.  
We adopt the convention that $\otimes$ binds more tightly than $(;\!)$.  
We elide all associativity and unit isomorphisms associated with monoidal categories,  and often omit subscripts on identities and natural transformations where it can be inferred.

@@

To encode equivalences between morphisms (or $\Sigma$-terms), the hom-sets of the corresponding free categories can be endowed with the additional structure of a semilattice induced by an idempotent, commutative and associative operator \textit{join} $+$ as presented in~\cite{ghica2024equivalencehypergraphsegraphsmonoidal}.
This endowment turns such categories into $\catname{SLat}$-enriched categories where $\catname{SLat}$ is a category of \textit{unbounded} semilattices and homomorphisms between them (see Appendix section~\ref{sec:appendix:slat}).

\begin{definition}[Semilattice-enriched category~\cite{Borceux_1994}]
    A semilattice-enriched category $\mathbb{C}$ is defined by the following data:
    \begin{itemize}
        \item a set of objects $\mathbb{C}$
        \item for every pair of objects $A,B \in \mathbb{C}$ --- a hom-object (or, hom-semilattice) $\mathbb{C}(A,B) \in \catname{SLat}$
        \item for every triple of objects $A,B,C \in \mathbb{C}$ --- a composition morphism
        \[
            \circ: \mathbb{C}(B,C) \otimes \mathbb{C}(A,B) \to \mathbb{C}(A,C)
        \]
        \item for every object $A \in C$ --- a unit morphism
        \[
        u_{A} : I \to \mathbb{C}(A,A), \text{ where $I$ is the monoidal unit for $\catname{SLat}$}
        \]
    \end{itemize}
    such that particular coherence diagrams commute.
\end{definition}

@@

    \begin{proposition}(A specialised case of Proposition 6.4.7~\cite{Borceux_1994})
    There is a 2-adjunction 
    % https://q.uiver.app/#q=WzAsMixbMCwwLCJcXGNhdG5hbWV7U0xhdHR9LVxcY2F0bmFtZXtDYXR9Il0sWzIsMCwiXFxjYXRuYW1le0NhdH0iXSxbMSwwLCJGIiwyLHsiY3VydmUiOjR9XSxbMCwxLCJHIiwyLHsiY3VydmUiOjR9XSxbMiwzLCIiLDAseyJsZXZlbCI6MSwic3R5bGUiOnsibmFtZSI6ImFkanVuY3Rpb24ifX1dXQ==
    \[\begin{tikzcd}
        \catname{SLat}\text{--}\catname{Cat}\arrow[rr, "\mathcal{U}", bend left] & \hspace{-1em}\top & \catname{Cat} \arrow[ll, "\mathcal{F}", bend left]
        \end{tikzcd}
    \]
    that is induced by a usual free-forgetful adjunction

    \[\begin{tikzcd}
        \catname{SLat} \arrow[rr, "U", bend left] & \top & \catname{Set} \arrow[ll, "F", bend left]
        \end{tikzcd}
    \]
    \end{proposition}

In particular, 2-functor $\mathcal{F}$ turns every category $\mathbb{C'} \in \catname{Cat}$ into a free semilattice enriched $\mathbb{C} \in \catname{SLat}\text{--}\catname{Cat}$ by making every hom-set of $\mathbb{C'}$ into a free semilattice on this set.
Being a 2-functor it also lifts an adjunction in $\catname{Cat}$ $F \dashv G$ into an $\catname{Slat}$-adjunction.
Thus, applying the functor to $\catname{CS}(\Sigma)$ we get a closed symmetric monoidal $\catname{Slat}$-category $\catname{CS}(\Sigma)^{+}$ by lifting the adjunction $(- \otimes B, B \multimap{} -, \eta, \varepsilon)$.
We can similarly derive $\catname{SLat}$-enriched versions of  $\Lambda_{A,B,C}$ and $\textbf{ev}_{A,B}$ morphisms using the lifted adjunction as $\textbf{ev}_{A,B} = \varepsilon_{A}$ and $\Lambda_{A,B,C}(f) = \eta; (B \multimap{} f)$.
The following distributivity law can then be derived, forming the basis of encoding equivalences inside abstractions.
\begin{equation}
\begin{aligned}
\Lambda_{A,B,C}(f + g) &= \eta;(B \multimap (f + g))\\
                       &= \eta;(B \multimap (f) + B \multimap (g))\\
                       &= \eta;(B \multimap (f)) + \eta;(B \multimap (g))\\
                       &= \Lambda_{A,B,C}(f) + \Lambda_{A,B,C}(g)
\end{aligned}
\end{equation}
\label{law:distributivity}

Similar to $\catname{CS}(\Sigma)$, $\catname{CS}(\Sigma)^{+}$ can be freely generated from a set of $\Sigma^{+}$-terms, the latter being defined as follows.
Every $\Sigma$-term is also a $\Sigma^{+}$-term. 
Given two $\Sigma^{+}$-terms $f : A \to B$ and $g : A \to B$, $f + g : A \to B$ is also a $\Sigma^{+}$-term.
Terms of $\catname{CS}(\Sigma)^{+}$ are quotiented by the laws of closed $\catname{SLat}$-SMC, including the distributivity law~\ref{law:distributivity}.

@@

In~\cite{ghica2024stringdiagramslambdacalculifunctional} it was shown that e-graphs over some signature $\Sigma$ can be represented as morphisms of a semilattice-enriched free Cartesian symmetric monoidal category over the same signature.
We argue that restricting the categorical domain to free semilattice-enriched closed symmetric monoidal categories gives rise to structures that naturally support binding and equivalence classes of morphisms.
That is, morphisms (string diagrams) of such category can both encode variable binding and equivalence between subterms (sub-diagrams).
We call these string diagrams, or more precisely, the combinatorial representation of that will follow, e-graphs with bindings.

\begin{figure*}
\[
\adjustbox{scale=0.75}{
\tikzfig{./figures/egraph-strings}
}
\]
\caption{String diagrams for closed semilattice-enriched symmetric monoidal categories.}
\label{fig:egraph-strings}
\end{figure*}

>>>>

Given a category $\mathbb{C}$  with objects $A,B \in \mathbb{C}$ we denote by $\mathbb{C}(A,B)$ the corresponding hom-set.  
We write the identity morphism on $A$ as $\id_A$.  
We commonly write $f;g$ for composition in diagrammatic order.  
Composition in the usual order is written $g \circ f$.  
We denote the tensor product of an SMC $\mathbb{C}$ by $\otimes$,  its unit by $I$, and its symmetry natural transformation as $\sym$ \cite{maclane}.  
We adopt the convention that $\otimes$ binds more tightly than $(;\!)$.  
We elide all associativity and unit isomorphisms associated with monoidal categories,  and often omit subscripts on identities and natural transformations where it can be inferred.
Given two adjoint functors $F : \mathbb{A} \to \mathbb{B}$ and $G : \mathbb{B} \to \mathbb{A}$ with $F$ being left-adjoint, we write $F \dashv G : \mathbb{B} \to \mathbb{A}$ for this adjunction.

@@

To encode equivalences between morphisms (or $\Sigma$-terms), the hom-sets of the corresponding free categories can be endowed with the additional structure of a semilattice induced by an idempotent, commutative and associative operator \textit{join} $+$ as presented in~\cite{ghica2024equivalencehypergraphsegraphsmonoidal}.
This endowment turns such categories into $\catname{SLat}$-enriched categories where $\catname{SLat}$ is a category of \textit{unbounded} semilattices and homomorphisms between them (see Appendix~\ref{sec:appendix:slat}).

\begin{definition}[Semilattice-enriched category~\cite{Borceux_1994}]
\label{def:slat-cat}
    A semilattice-enriched category $\mathbb{C}$ is defined by the following data:
    \begin{itemize}
        \item a set of objects $\mathbb{C}$
        \item for every pair of objects $A,B \in \mathbb{C}$ --- a hom-object (or, hom-semilattice) $\mathbb{C}(A,B) \in \catname{SLat}$
        \item for every triple of objects $A,B,C \in \mathbb{C}$ --- a composition morphism
        \[
            \circ: \mathbb{C}(B,C) \otimes \mathbb{C}(A,B) \to \mathbb{C}(A,C)
        \]
        \item for every object $A \in C$ --- a unit morphism
        \[
        u_{A} : I \to \mathbb{C}(A,A), \text{ where $I$ is the monoidal unit for $\catname{SLat}$}
        \]
    \end{itemize}
    such that particular coherence diagrams commute.
\end{definition}
As the composition is a semilattice homomorphism (morphism of $\catname{SLat}$), we have the following equations 
\[
(f + g); h = f;h + g;h \quad f;(g+h) = f;g + f;h
\]

@@

    Then, the axioms of $\catname{SLat}$-SMC consist of equations in Definitions~\ref{def:enriched-prop}~\ref{def:slat-cat} and associativity, commutativity, and idempotence equations for $+$.

    \begin{proposition}(A specialised case of Proposition 6.4.7~\cite{Borceux_1994})
        There is a 2-adjunction 
        % https://q.uiver.app/#q=WzAsMixbMCwwLCJcXGNhdG5hbWV7U0xhdHR9LVxcY2F0bmFtZXtDYXR9Il0sWzIsMCwiXFxjYXRuYW1le0NhdH0iXSxbMSwwLCJGIiwyLHsiY3VydmUiOjR9XSxbMCwxLCJHIiwyLHsiY3VydmUiOjR9XSxbMiwzLCIiLDAseyJsZXZlbCI6MSwic3R5bGUiOnsibmFtZSI6ImFkanVuY3Rpb24ifX1dXQ==
        % \[\begin{tikzcd}
        % 	\catname{SLatt}\text{--}\catname{Cat}\arrow[rr, "\mathcal{U}", bend left] & \hspace{-1em}\top & \catname{Cat} \arrow[ll, "\mathcal{F}", bend left]
        % 	\end{tikzcd}
        % \]
        \vspace{-1mm}
        \[
        \mathcal{F} \dashv \mathcal{U} : \catname{SLat}\text{--}\catname{Cat} \to \catname{Cat}
        \vspace{-1mm}
        \]
        that is induced by a usual free-forgetful adjunction

        % \[\begin{tikzcd}
        % 	\catname{SLatt} \arrow[rr, "U", bend left] & \top & \catname{Set} \arrow[ll, "F", bend left]
        % 	\end{tikzcd}
        % \]
        \vspace{-1mm}
        \[
        F \dashv U : \catname{SLat} \to \catname{Set}
        \vspace{-1mm}
        \]        
    \end{proposition}
In particular, 2-functor $\mathcal{F}$ turns every category $\mathbb{C'} \in \catname{Cat}$ into a free semilattice enriched $\mathbb{C} \in \catname{SLat}\text{--}\catname{Cat}$ by making every hom-set of $\mathbb{C'}$ into a free semilattice on this set.
Being a 2-functor, it also lifts an adjunction in $\catname{Cat}$ $F \dashv G$ into an $\catname{Slat}$-adjunction.
Thus, applying the functor to $\catname{CS}(\Sigma)$ we get a closed symmetric monoidal $\catname{Slat}$-category $\catname{CS}(\Sigma)^{+}$ by lifting the adjunction $(- \otimes B, B \multimap{} -, \eta, \varepsilon)$.
We can similarly derive $\catname{SLat}$-enriched versions of  $\Lambda_{A,B,C}$ and $\textbf{ev}_{A,B}$ morphisms using the lifted adjunction as $\textbf{ev}_{A,B} = \varepsilon_{A}$ and $\Lambda_{A,B,C}(f) = \eta; (B \multimap{} f)$.
The following distributivity law can then be derived, forming the basis of encoding equivalences inside abstractions.
\begin{equation}
\begin{aligned}
\Lambda_{A,B,C}(f + g) &= \eta;(B \multimap (f + g))\\
                       &= \eta;(B \multimap (f) + B \multimap (g))\\
                       &= \eta;(B \multimap (f)) + \eta;(B \multimap (g))\\
                       &= \Lambda_{A,B,C}(f) + \Lambda_{A,B,C}(g)
\end{aligned}
\end{equation}
\label{law:distributivity}

Similar to $\catname{CS}(\Sigma)$, $\catname{CS}(\Sigma)^{+}$ can be freely generated from a set of closed $\Sigma^{+}$-terms, the latter being defined as follows.
Every closed $\Sigma$-term is also a $\Sigma^{+}$-term. 
Given two closed $\Sigma^{+}$-terms $f : A \to B$ and $g : A \to B$, $f + g : A \to B$ is also a closed $\Sigma^{+}$-term.
Terms of $\catname{CS}(\Sigma)^{+}$ are quotiented by the laws of closed $\catname{SLat}$-SMC, including the distributivity law~\ref{law:distributivity}.

@@

% In~\cite{ghica2024stringdiagramslambdacalculifunctional} it was shown that e-graphs over some signature $\Sigma$ can be represented as morphisms of a semilattice-enriched free Cartesian symmetric monoidal category over the same signature.
% We argue that restricting the categorical domain to free semilattice-enriched closed symmetric monoidal categories gives rise to structures that naturally support binding and equivalence classes of morphisms.
% That is, morphisms (string diagrams) of such category can both encode variable binding and equivalence between subterms (sub-diagrams).
% We call these string diagrams, or more precisely, the combinatorial representation of that will follow, e-graphs with bindings.

\begin{figure*}
\[
\adjustbox{scale=0.65}{
\tikzfig{./figures/egraph-strings}
}
\]
\captionsetup{skip=0pt, belowskip=-2ex}
\caption{String diagrams for closed semilattice-enriched symmetric monoidal categories.}
\label{fig:egraph-strings}
\end{figure*}
